Concurrent Programming in Harmony

### 2021-12-26 book/paper.tex
Rewrote the Chain Replication chapter

\begin{figure}
\begin{code}
<{:input:../code/chain.hny}>
\end{code}
\caption{\harmonylink{code/chain.hny} Checking correctness of Chain Replication}
\label{fig:chain}
\end{figure}

@@

\glsadd{state machine replication}%
\index{state machine replication}%
\index{determinism}%
Besides availability, it is usually important that the replicated
service acts as if it were a single one.  This requires that the
replicas of the service coordinate their actions.
The \emph{Replicated State Machine Approach}~\cite{Lam78,S90} is a
general approach to do just this.
First, you model your service as a deterministic state
machine.  The replicas each run a copy of the state machine, started
in the same state.  As long as the replicas handle the same inputs
in the same order, determinism guarantees that they produce the same
outputs in the same order.
Therefore, the trick is to ensure that all replicas handle the
same requests in the same order and to do so in a way that continues
to work even if some strict subset of replicas crash.

\emph{Chain Replication}~\cite{vRS04} is such a replication protocol.
In Chain Replication, the replicas are organized in a linear chain.
Each replica monitors its direct predecessor in the chain.  If a replica has
no predecessors, we call it the \emph{head}.  The head may change over
time, as a head may crash and replaced by another, but at any point in
time there is only one head
(and only one \emph{tail}, possibly the same replica).

We model the state machine as a \emph{history}: a sequence of operations.
Only the head is allowed to introduce new updates.
When it does so, it advertises the new update along with its position
in the history.  When the direct successor receives the next update
it is expecting, it appends the update to its history and likewise
advertises the new update.  When the update is applied to the history
of the tail replica (the replica that has no live successors),
then the update is considered final.

So, when a replica fails, its successors should find out about it.
In practice, one server can detect the failure of another server by
pinging it.  If a server does not receive a response to its ping within
some maximum amount of time, then the server considers its peer crashed.
Note that this, in general, is not a safe thing to do---the network or
the peer may be temporarily slow but the peer is not necessarily crashed
when the timer expires.
Nonetheless, we will assume here that failure detection does not make
mistakes and that eventually every failure is eventually detected.
This is called the \emph{Fail-Stop} failure model~\cite{SS83},
which is distinct from the often more realistic \emph{Crash} failure model
where processes can crash but accurate detection is not available.
We will consider that failure model in the upcoming chapters.

\autoref{fig:chain} shows an implemenation of chain replication.
The code starts \texttt{NREPLICAS} \texttt{replica} threads numbered
0 through $\mathtt{NREPLICAS} - 1$, with initially 0 being the head
and $\mathtt{NREPLICAS} - 1$ being the tail.  At least one of the replicas 
must be immortal---the code selects one of them although none of the
replicas is required to fail during execution.
Of course, a replica does not know whether it is immortal
or not in practice---it should just assume that it is.
The immortality of one of the replicas is only used for modeling
the assumptions we make about the system.

Each replica maintains its history and a set of its predecessors.
It then loops until either it fails or it has applied all
<{NOPS}> operations to its history.
<{choose({ immortal, True })}> can only
evaluate to <{False}> in case <{immortal = False}>; otherwise
it will always evaluate to <{True}>.
What it does within the loop depends on whether it is the head
(has no alive predecessors) or not.
Because failure detection is accurate, at most one replica can think
it is the head at any time (and, if so, it is in fact the head).
Moreover, when it has
detected all its predecessors having failed, eventually some replica
thinks it is the head.

Messages have the format $(\mathit{source}, \mathit{payload})$,
where <{source}> is the identifier of the replica that generated
the message and <{payload}> has information about the state of
the replica.  The payload can take one of two forms.  The first is
<{.crash}>, indicating that the source of the message has
crashed.  The code models a replica crashing by that replica broadcasting
a <{.crash}> message on the network.
The other form of payload is a tuple
$(\mathit{index}, \mathit{operation})$, which means that the source
placed the given operation at the given index in its history.

If a replica thinks it is the head, it adds an operation to its
history.
In practice this would be some operation that the head would have received
from a client.  Clients send their operations to the head, which adds them
to its history in the order received.
We are not modeling clients of the service.
Instead, in the model, the head adds its identifier <{self}> to the
history until the
history has <{NOPS}> operations on it or until the head crashes,
whichever comes first.  The head also broadcasts each update, which
are intended for its successor.  Because of failures, it may be that
at different times a replica has different successors, hence the
broadcast.

Otherwise, the replica does not believe it is the head and
awaits a message from its direct predecessor.
If the payload is <{.crash}>, then the replica removes this
predecessor from its set of predecessors.  For example, if there
are initially three replicas numbered 0, 1, and 2, and replica~1
crashes, then replica~2 removes~1 from its set of predecessors, making
replica~0 its direct (and only) predecessor.  If, instead, the payload
is a tuple, then it must be the case that it contains the next operation
to add to its history.  The replica adds the operation to its history
and broadcasts the update, again intended for its successor.

When the replica successfully adds all <{NOPS}> updates to its
history without crashing, the code makes sure that any other such replica
ends in the same state.  Variable <{final}> is used to check the
different histories against one another.

>>>>

\begin{figure}
\begin{code}
<{:input:../code/rsmspec.hny}>
\end{code}
\caption{\harmonylink{code/rsmspec.hny} Replicated State Machine}
\label{fig:rsmspec}
\end{figure}

\begin{figure}
\begin{code}
<{:input:../code/chain.hny}>
\end{code}
\caption{\harmonylink{code/chain.hny} Chain Replication}
\label{fig:chain}
\end{figure}

@@

\glsadd{state machine replication}%
\index{state machine replication}%
\index{determinism}%
Besides availability, it is usually important that the replicated
service acts as if it were a single one.  This requires that the
replicas of the service coordinate their actions.
The \emph{Replicated State Machine Approach}~\cite{Lam78,S90} is a
general approach to do just this.
First, you model your service as a deterministic state
machine.  The replicas each run a copy of the state machine, started
in the same state.  As long as the replicas handle the same inputs
in the same order, determinism guarantees that they produce the same
outputs in the same order.

\autoref{fig:rsmspec} presents a Harmony specification of state machine
replication.
We model the state machine as a \emph{history}: a sequence of operations.
Each replica has a local copy of the history.
The network is modeled as an ordered queue of messages.
<{NOPS}> clients each place an operation on the network.
All but one of the replicas is allowed to crash.
The model chooses one of the replicas that is not allowed to crash.
Of course, a replica does not know whether it is immortal
or not in practice---it should just assume that it is.
The immortality of one of the replicas is only used for modeling
the assumptions we make about the system.
A replica loops until either it fails or it has received all operations.
<{choose({ immortal, True })}> can only
evaluate to <{False}> in case <{immortal = False}>; otherwise
it will always evaluate to <{True}>.
After a replica has received all <{NOPS}> operations, the replica
finishes and prints its history.
All replicas will print the same history (which should be the same
as the network).

But in reality the network is not an ordered queue and better modeled
as a set of messages.
The trick now is to ensure that all replicas handle the
same requests in the same order and to do so in a way that continues
to work even if some strict subset of replicas crash.
\emph{Chain Replication}~\cite{vRS04} is such a replication protocol.
In Chain Replication, the replicas are organized in a linear chain.
Each replica monitors its direct predecessor in the chain.  If a replica has
no predecessors, we call it the \emph{head}.  The head may change over
time, as a head may crash and replaced by another, but at any point in
time there is only one head
(and only one \emph{tail}, possibly the same replica).

Only the head is allowed to accept new updates from clients.
When it does so, it adds the update to the end of its history
and sends the history to its successor on the chain.
When the direct successor receives such a history, it makes sure
that the history is an extension of its own and, if so, replaces its own
history with the one received.
It then sends the history on to its successor, if any.

So, when a replica fails, its successors should find out about it.
In practice, one server can detect the failure of another server by
pinging it.  If a server does not receive a response to its ping within
some maximum amount of time, then the server considers its peer crashed.
Note that this, in general, is not a safe thing to do---the network or
the peer may be temporarily slow but the peer is not necessarily crashed
when the timer expires.
Nonetheless, we will assume here that failure detection does not make
mistakes and that eventually every failure is eventually detected.
This is called the \emph{Fail-Stop} failure model~\cite{SS83},
which is distinct from the often more realistic \emph{Crash Failure} model
where processes can crash but accurate detection is not available.
We will consider that more realistic failure model in the upcoming chapters.

\autoref{fig:chain} shows an implemenation of chain replication.
Each replica maintains its history and a set of its predecessors.
What it does within the loop depends on whether it is the head
(has no alive predecessors) or not.
Because failure detection is accurate, at most one replica can think
it is the head at any time (and, if so, it is in fact the head).
Moreover, when it has
detected all its predecessors having failed, eventually some replica
thinks it is the head.

Messages have the format $(\mathit{source}, \mathit{payload})$,
where <{source}> is the identifier of the replica that generated
the message and <{payload}> has information about the state of
the replica.  The payload can take one of two forms.  The first is
<{.crash}>, indicating that the source of the message has
crashed.  The code models a replica crashing by that replica broadcasting
a <{.crash}> message on the network.
The other form is a history of operations.

If a replica thinks it is the head, it adds an operation to its
history.  The head then broadcasts its history, which
is intended for its successor.  Because of failures, it may be that
at different times a replica has different successors, hence the
broadcast.

Otherwise, the replica does not believe it is the head and
awaits a message from its direct predecessor.
If the payload is <{.crash}>, then the replica removes this
predecessor from its set of predecessors.  For example, if there
are initially three replicas numbered 0, 1, and 2, and replica~1
crashes, then replica~2 removes~1 from its set of predecessors, making
replica~0 its direct (and only) predecessor.
Otherwise, the payload must be a history.
The replica only accepts the history if it is an extension of its own
history.
When the replica successfully received all <{NOPS}> updates
without crashing, it prints its history.

The question is whether \autoref{fig:chain} has the same behavior as
the replicated state machine specification of \autoref{fig:rsmspec}.
This can be checked using the following two Harmony commands:
\begin{code}
\begin{verbatim}
$ harmony -o rsm.hfa code/rsmspec.hny
$ harmony -B rsm.hfa code/chain.hny
\end{verbatim}
\end{code}

The first command outputs the behavior DFA of \texttt{code/rsmspec.hny}
in the file \texttt{rsm.hfa}.
The second command checks that the behavior of \texttt{code/chain.hny}
satisfies the DFA in \texttt{rsm.hfa}.
The DFA can also be visualized using the command
\texttt{harmony -o rsm.png code/rsmspec.hny}
(\autoref{fig:rsmdfa}).
It shows all permutations of the operations are possible outputs.
Moreover, each is output at least once, and at most
<{NREPLICAS}> times.

\begin{figure}
\begin{center}
\includegraphics[width=0.8\textwidth]{figures/rsmspec.png}
\end{center}
\caption{The DFA generated by both
\autoref{fig:rsmspec} and \autoref{fig:chain}}
\label{fig:rsmdfa}
\end{figure}
